Adding selection to \stdpopsim: oh how the tables have turned

### 2024-11-18 analysis2_manuscript.tex
edits to other sections

\section*{Abstract}
Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
However, simulating realistic models that incorporate both selection and complex demographic histories
has remained challenging, limiting our ability to benchmark statistical methods and explore theoretical predictions.
Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
for species-specific genomic annotations and published DFE estimates. 
We demonstrate the utility of this framework by benchmarking methods for demographic inference,
DFE estimation, and selective sweep detection across different species and scenarios. 
Our results highlight how selection can bias demographic inference, 
reveal the sensitivity of DFE inference methods to model assumptions, 
and show how genomic features like recombination rate and functional density influence power to detect selective sweeps.
This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
to explore the interplay between selection and other evolutionary forces in a reproducible framework.

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \cite[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties, barton1998effect},
    % DRS -- I'm not sure Barton reference is right for this?
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \cite{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However this is a challenging task from a modeling perspective \citep{johri2022prospect}.

    % simulation
    % DRS: lots more refs will be needed in this paragraph. added a few and am happy to help add more later if needed
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting}, to
    to explore the power and limitations of statistical methods \cite[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, as well as a number of parameter choices including
    the strength of selection, the distribution of fitness effects, and the
    recombination rate, as well as a parameterized model of demography.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, sharing and comparing simulation results among studies can 
    itself be challenging, as different simulation tools may use different structures
    and parameter scalins for their models, and may not be directly comparable.
    This is a major limitation for the field, as it can make it difficult
    to assess the robustness of results, and to compare the results of
    different studies.

    % stdpopsim
    % DRS: lots more refs will be needed in this paragraph. added a few and am happy to help add more later if needed
    A lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirical
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features after estimates/annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background}.

@@

    % current numbers of DFEs / species / etc
    At this release the \stdpopsim catalog comprises 24 species (i.e. genome representations)
    for which we have implemented XX demographic models, have XX genetic maps, and XX DFEs.
    Additions to the catalog are ongoing -- we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding} and we welcome contributions from the
    community---we encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

@@

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the pairwise sequentially Markovian coalescent
    (PSMC) \citep{li2011inference}, through use of the site frequency spectrum (SFS),
    as well as through identity by decent information. 
    While that is so, it has been well described that selection, because of its effects on
    local and more global patterns of genetic variation, can drive estimates of $N_e(t)$
    further from true census population sizes simply because selection increases the 
    rate of coalescence in the genome.

    Using \stdpopsim we can simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography} we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without background selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome. 
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we have performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted compute as thorough benchmarking
    isn't the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography} we show six panels, each of which shows the estimated $N_e(t)$
    from each independent sample in a five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of simulations without selection, while the bottom row shows the results
    of simulations with a Gamma distributed DFE for the deleterious effects of mutations occurring exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the bottom row are to visually reinforce that inference is done
    from a model with selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ in the more recent past for the CEU and CHB populations.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwaiplot which uses the SFS to infer $N_e(t)$ seems to slightly underestimate $N_e(t)$
    but is reasonably close to the true $N_e(t)$ in the absence of selection.
    Likewise \smcpp seems to be more robust to the presence of selection,
    but there is some added noise to the returned estimates. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$.
    % DRS: I don't agree with the interpretation in the previous two sentences (starting with "Likewise").
    % MSMC is already barfing a little bit even without selection, and it is not clear to me from these
    % plots that selection makes things any worse. Also, SMC++ is doing quite well without selection so maybe
    % the minor drop in accuracy is meaningful in relative terms.

@@

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data-- 
    \dadi \citep{gutenkunst2009inferring}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadi seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods slightly underestimated the strength of selection. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadi seems to yield overestimates.

@@

    In Figures \ref{fig:vaquita-dfe.constant} and \ref{fig:vaquita-dfe} we show a comparison of estimates
    from the same three software packages for inferring the DFE from genetic data, but for simulations of the
    vaquita porpoise genome. In the constant size model (fig \ref{fig:vaquita-dfe.constant}) we see that all methods
    perform uniformly worse that in the human genome simulations, which consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter. 
A critical distinction is that the previous human simulations employed a DFE that assumed all selected mutations had additive effects (no dominance, $h = 1/2$), whereas the vaquita simulations use a DFE in which more deleterious mutations are more recessive ($h < 1/2$).
For all three software packages, the inferred mean and shape of the distribution of selection coefficients is consistent with the mean and shape of the distribution of $2 h s$, the product of the dominance coefficient and the selection coefficient (Fig.~\ref{fig:vaquita-dfe}C).
This is likely because deleterious alleles are typically at low frequency and thus heterozygous, where their selective effect is $h s$, and all these tools assume additivity  ($h = 1/2$).
Strongly deleterious alleles are often observed to be recessive in many species \citep{}, so these results suggest caution in interpreting DFE inferences.
%RNG: Could use Kirk's help with the best references here

@@

\section*{Funding}
    \label{funding}

\printbibliography

>>>>

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulating realistic models that incorporate both selection and complex demographic histories
    has remained challenging, limiting our ability to benchmark statistical methods and explore theoretical predictions.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
    This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across different species and scenarios. 
    Our results highlight how selection can bias demographic inference, 
    reveal the sensitivity of DFE inference methods to model assumptions, 
    and show how genomic features like recombination rate and functional density influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible framework.

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \cite[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \cite{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However this is a challenging task from a modeling perspective \citep{johri2022prospect}.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting}, to
    to explore the power and limitations of statistical methods \cite[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, as well as a number of parameter choices including
    the strength of selection, the distribution of fitness effects, and the
    recombination rate, as well as a parameterized model of demography.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, sharing and comparing simulation results among studies can 
    itself be challenging, as different simulation tools may use different structures
    and parameter scalins for their models, and may not be directly comparable.
    This is a major limitation for the field, as it can make it difficult
    to assess the robustness of results, and to compare the results of
    different studies.

    % stdpopsim
    A lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirical
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features after estimates/annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background}.

@@

    % current numbers of DFEs / species / etc
    At this release the \stdpopsim catalog comprises 24 species (i.e. genome representations)
    for which we have implemented 28 demographic models, have 37 genetic maps, and 7 DFEs. %someone should double check the numbers
    Additions to the catalog are ongoing -- we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding} and we welcome contributions from the
    community---we encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

@@

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the pairwise sequentially Markovian coalescent
    (PSMC) \citep{li2011inference}, through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by decent information \citep{santiago2020recent}. 
    While that is so, it has been well described that selection, because of its effects on
    local and more global patterns of genetic variation, can drive estimates of $N_e(t)$
    further from true census population sizes simply because selection increases the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}.

    Using \stdpopsim we can simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography} we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without background selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome. 
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we have performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted computational resources as thorough benchmarking
    isn't the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography} we show six panels, each of which shows the estimated $N_e(t)$
    from each independent sample in a five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of simulations without selection, while the bottom row shows the results
    of simulations with a Gamma distributed DFE for the deleterious effects of mutations occurring exons.
    In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the bottom row are to visually reinforce that inference is done
    from a model with selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ in the more recent past for the CEU and CHB populations.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwaiplot which uses the SFS to infer $N_e(t)$ seems to slightly underestimate $N_e(t)$
    but is reasonably close to the true $N_e(t)$ in the absence of selection.
    Likewise \smcpp seems to be more robust to the presence of selection,
    but there is some added noise to the returned estimates. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$,
    although estimates from \msmc have quite a bit of added noise even in the absence of selection.

    % DRS: I don't agree with the interpretation in the previous two sentences (starting with "Likewise").
    % MSMC is already barfing a little bit even without selection, and it is not clear to me from these
    % plots that selection makes things any worse. Also, SMC++ is doing quite well without selection so maybe
    % the minor drop in accuracy is meaningful in relative terms.
    % ADK: I edited the text-- what do you think Dan?

@@

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data-- 
    \dadi \citep{gutenkunst2009inferring}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as a current limitation of \stdpopsim
    currently is that we are not able to simulate an outgroup sequence to estimate
    divergence. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadi seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods slightly underestimated the strength of selection. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadi seems to yield overestimates.

@@

    In Figures \ref{fig:vaquita-dfe.constant} and \ref{fig:vaquita-dfe} we show a comparison of estimates
    from the same three software packages for inferring the DFE from genetic data, but for simulations of the
    vaquita porpoise genome. In the constant size model (fig \ref{fig:vaquita-dfe.constant}) we see that all methods
    perform uniformly worse that in the human genome simulations, with consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter. 
    A critical distinction is that the previous human simulations employed a DFE that assumed all selected mutations had additive effects
    (no dominance, $h = 1/2$), whereas the vaquita simulations use a DFE in which more deleterious mutations are more recessive ($h < 1/2$).
    For all three software packages, the inferred mean and shape of the distribution of selection coefficients 
    is consistent with the mean and shape of the distribution of $2 h s$, the product of the dominance coefficient
    and the selection coefficient (Fig.~\ref{fig:vaquita-dfe}C).
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous, 
    where their selective effect is $h s$, and all these tools assume additivity  ($h = 1/2$).
    Strongly deleterious alleles are often observed to be recessive in many species \citep{mukai1972mutation},
    so these results suggest caution in interpreting DFE inferences.
    %RNG: Could use Kirk's help with the best references here

@@

\section*{Funding}
    \label{funding}
    Multiple funding sources supported this work.
    MFR, ST, NP, and ADK were supported in part by NIH award R35GM148253.
    %other funding sources?
\printbibliography
